---
title: "Weight-for-height"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Weight-for-height}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Note that weight for height calculations use `x_var` as 'htcm' or 'lencm' and that the current functions do not account for the age category of the coefficients tables by using the age variable in your data.  You are expected to subset your data to the 0-2 age category and the use 'lencm' as your `x_var` to get the appropriate calculations.  Similarly, if you use 'htcm' as the `x_var` then the 2-5 age category will be used for WHZ calculations.


```{r, echo=TRUE, eval=TRUE}
library(growthstandards)
```

## dplyr method

With `dplyr` the use of `case_when()` we can calculate 'whz' using 'agedays' to create our two cases. 

```{r, eval=FALSE}
library(dplyr)
cpp %>%
  mutate(
    whz = case_when(
      agedays < 730 ~ who_value2zscore(
        x = htcm, y = wtkg, x_var = "lencm", y_var = "wtkg", sex = sex),
      agedays >= 730 ~ who_value2zscore(
        x = htcm, y = wtkg, x_var = "htcm", y_var = "wtkg", sex = sex)
    )
  )
```

## Base R method

Using the provided `cpp` data we have created two data sets based on the 'agedays' variable. 


```{r}
cpp_lt2 <- subset(cpp, agedays < 730)
cpp_gt2 <- subset(cpp, agedays >= 730)
```

Now we can use the specific functions to calculate the appropriate z-scores.

```{r}

cpp_lt2$whz <- who_value2zscore(
  x = cpp_lt2$htcm, y = cpp_lt2$wtkg, x_var = "lencm", y_var = "wtkg", 
  sex = cpp_lt2$sex)

cpp_gt2$whz <- who_value2zscore(
  x = cpp_gt2$htcm, y = cpp_gt2$wtkg, x_var = "htcm", y_var = "wtkg", 
  sex = cpp_gt2$sex)

```

Finally, we can combine the data back into one data set with the additional 'whz' column.

```{r}
cpp_whz <- rbind(cpp_lt2, cpp_gt2)
cpp_whz <- cpp_whz[order(cpp_whz$subjid, cpp_whz$agedays), ]
```



